---
title: "Multigroup transmission and vaccination model"
subtitle: "A tool for investigating intervention impacts on health disparities"
author: Damon Toth
institute: University of Utah
format:
  revealjs:
    footer: "Damon Toth"
    self-contained-math: true
    theme: simple
    slide-number: c
---

## Motivation {.smaller}

-   Infectious disease outbreaks can have disparate impacts across different population sub-groups

-   Higher infection rates in one group can occur by several mechanisms

    -   Higher susceptibility to the infectious organism

    -   Higher overall rates of person-to-person contact

    -   Transmission amplification due to higher within-group contact

-   Disparities in vaccination coverage can also exist

    -   How do efforts to reduce vaccine disparity affect transmission outcomes?

    -   Can interventions aimed at reducing disparities be cost effective?

## Two-group mathematical model {.smaller}

Susceptible ($S$) -- Infectious ($I$) -- Removed ($R$) model with two groups (A and B)

::::::: columns
:::: {.column width="50%"}
::: {style="font-size: 70%;"}
$$\frac{dS_A}{dt} = -\left(\beta_{AA}\frac{I_A}{N_A} + \beta_{AB}\frac{I_B}{N_B}\right)S_A
$$ $$\frac{dS_B}{dt} = -\left(\beta_{BA}\frac{I_A}{N_A} + \beta_{BB}\frac{I_B}{N_B}\right)S_B
$$ $$\frac{dI_A}{dt} = \left(\beta_{AA}\frac{I_A}{N_A} + \beta_{AB}\frac{I_B}{N_B}\right)S_A - \gamma I_A
$$ $$\frac{dI_B}{dt} = \left(\beta_{BA}\frac{I_A}{N_A} + \beta_{BB}\frac{I_B}{N_B}\right)S_B - \gamma I_B
$$ $$\frac{dR_A}{dt} = \gamma I_A
$$ $$\frac{dR_B}{dt} = \gamma I_B
$$ Vaccination: instantaneously remove portion of susceptibles
:::
::::

:::: {.column width="50%"}
::: {style="font-size: 70%;"}
-   $N_A$, $N_B$: Group population sizes
-   $\gamma$: End-of-infectiousness rate
-   $\beta_{ij}$: Transmission rates:

$$
\beta_{ii} = q_ia_i\left(\epsilon_i+(1-\epsilon_i) \frac{(1-\epsilon_i)a_iN_i}{(1-\epsilon_i)a_iN_i+(1-\epsilon_j)a_jN_j}\right)
$$ $$
\beta_{ij} = q_ia_i(1-\epsilon_i) \frac{(1-\epsilon_j)a_jN_j}{(1-\epsilon_i)a_iN_i+(1-\epsilon_j)a_jN_j},\quad i\neq j
$$

-   $q_i$: probability of infection per at-risk contact for group $i$

-   $a_i$: overall contact rate of group $i$

-   $\epsilon_i$: portion of group $i$'s contacts exclusively within group

    -   $\epsilon_i=0$: contacts occur proportionally to overall group sizes and contact rates

    -   $\epsilon_i=1$: all contacts are within own group
:::
::::
:::::::

## Input components {.smaller}

-   Transmission model

    -   Population sizes ($N_A$, $N_B$)

    -   Overall transmissibility of pathogen (Basic reproduction number $R_0$)

    -   Relative susceptibility per contact ($q_B/q_A$)

    -   Relative overall contact rate ($a_A/a_B$)

    -   Within-group preferential mixing levels ($\epsilon_A$, $\epsilon_B$)

    -   End-of-infectiousness rate ($\gamma$)

-   Vaccination model

    -   Time at which all susceptibles are immunized

    -   Portion of each group immunized

## Final outbreak size calculation {.smaller}

With the input components from the previous slide, the `getFinalSize()` function does the following:

-   Computes the $\beta_{ij}$ transmission rate values

-   Simulates the differential equations up to the vaccination time (if \>0)

-   Calculates (simulation-free) the final size of the outbreak for each group

```{r, echo=TRUE, output=FALSE}
library(vaccine.equity)
finalsize <- getFinalSize(vacTime=0, vacPortion=c(0.1,0.1), popSize = c(800000, 200000), R0 = 1.5,
                          recoveryRate = 1/7, contactRatio = 1.7, contactWithinGroup = c(0.4, 0.4),
                          suscRatio = 1)
```
```{r}
print(round(finalsize))
```
